---
title: 'Reproducible Research: Peer Assessment 1'
output:
html_document:
keep_md: yes
pdf_document: default
word_document: default
---

## Loading and preprocessing the data
```{r load_and_preprocess, message=FALSE, warning=FALSE}
# loading all required libraries
library(tools)
library(plyr)
library(ggplot2)
library(gridExtra)


# file informationn
filelink <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
filename <- "activity.zip"
filechecksum <- "61e0c1f8c4a736ff1ab0f8863344b301"

# check to see if file is missing or checksum has changed 
if (!file.exists(filename)|md5sum(filename) != filechecksum) {
        # download the file.
        download.file(filelink, filename)
        }         

# unzip and read in CSV file and remove incomplete cases
unzip("activity.zip")
activity <- read.csv("activity.csv")
```
The data set that was used for the analysis is the one that was included in the [RepData_PeerAssessment1 Git repository](https://github.com/rdpeng/RepData_PeerAssessment1) (commit 80edf39c3bb508fee88e3394542f967dd3fd3270).  It can also be downloaded from [`r filelink`](`r filelink`).

## What is mean total number of steps taken per day?
The following histogram shows the distribution of frequencies of the number of steps taken each day:
```{r dialy_steps, message=FALSE}
# summarize the data by date
daily_steps <- ddply(activity,
                     c("date"),
                     summarise,
                     total_steps = sum(steps,
                                       na.rm = TRUE)
                     )

# create histogram
ggplot(daily_steps, 
       aes(x=total_steps)
       ) +
        geom_histogram(colour = "black",
                       fill = "blue"
                       )+ 
        labs(title = "Histogram of Total Daily Steps", 
             y = "Frequency",
             x = "Total Daily Steps"
             )

```

```{r dialy_steps_details, message=FALSE, echo=FALSE}
mean_daily_steps <- prettyNum(round(mean(daily_steps$total_steps), 0), big.mark = ",")
median_daily_steps <- prettyNum(median(daily_steps$total_steps), big.mark = ",")
```

The average number of steps taken was `r mean_daily_steps` and the median was `r median_daily_steps`.

## What is the average daily activity pattern?
The following chart
```{r activity_pattern}
# simmarize the data by interval
activity_pattern <- ddply(activity,
                          c("interval"),
                          summarize,
                          average_steps = mean(steps,
                                               na.rm = TRUE
                                               )
                          )

# create time series plot by interval
ggplot(activity_pattern,
       aes(x = interval,
           y = average_steps
           )
       ) +
        geom_line(colour = "blue",
                  size = 1
                  ) + 
        labs(title = "Average Steps per Interval", 
             y = "Average Steps",
             x = "Interval"
             ) 



# find peak interval
max_steps <- activity_pattern[with(activity_pattern, order(-average_steps)),][1,]
max_steps
```

```{r activity_pattern_details, echo=FALSE}
max_interval <- prettyNum(max_steps$interval[1], big.mark = ",")
max_avg <- prettyNum(round(max_steps$average_steps[1], 0), big.mark = ",")
```


The interval with the highest average number of steps was `r max_interval` with an average of `r max_avg` steps.

## Imputing missing values


```{r Imputing_values, message=FALSE}
# get the median number of steps per interval
msbi <- ddply(activity,
              c("interval"),
              summarize,
              median_steps = median(steps,
                                    na.rm = TRUE
                                    )
              )

# create a new data frame replacing NA values with the meidan value for that interval
activity_2 <- data.frame(interval = activity$interval,
                         date = activity$date, 
                         steps=ifelse(is.na(activity$steps), 
                                      msbi[match(activity$interval, msbi$interval), 2],
                                      activity$steps
                                      ), 
                         steps_old = activity$steps
                         )

# summarize the data by date
daily_steps_2 <- subset(ddply(activity_2, 
                              c("date"),
                              summarize, 
                              total_steps = sum(steps,
                                                na.rm = TRUE
                                                )
                              ),
                        total_steps != 0
                        )

# Generate histogram
ggplot(daily_steps_2, 
       aes(x=total_steps)
       ) +
        geom_histogram(colour = "black",
                       fill = "blue"
                       )+ 
        labs(title = "Histogram of Total Daily Steps", 
             y = "Frequency",
             x = "Total Daily Steps"
             )

```

```{r Imputing_values_details, echo=FALSE}
mean_daily_steps_2 <- prettyNum(round(mean(daily_steps_2$total_steps), 0), big.mark = ",")
median_daily_steps_2 <- prettyNum(median(daily_steps_2$total_steps), big.mark = ",")
mean_diff <- prettyNum(round(mean(daily_steps_2$total_steps) - mean(daily_steps$total_steps), 0), big.mark = ",")
median_diff <- prettyNum(median(daily_steps_2$total_steps) - median(daily_steps$total_steps), big.mark = ",")
```
The average number of steps taken was `r mean_daily_steps_2` (a change of `r mean_diff`) and the median was `r median_daily_steps_2` (a change of `r median_diff`).

## Are there differences in activity patterns between weekdays and weekends?


```{r weekday_vs_weekend, message=FALSE}
# get day of week from date and check to see if it's a weekend or weekday
activity_2$daytype <- ifelse(weekdays(as.Date(activity_2$date)) %in% c("Saturday","Sunday"),
                             "Weekend", 
                             "Weekday"
                             )
# simmarize the data by interval and day type
activity_pattern_2 <- ddply(activity_2,
                            c("daytype", "interval"),
                            summarize,
                            average_steps = mean(steps,
                                                 na.rm = TRUE
                                                 )
                            )

# generate weekday and weekend timeseries plots by interval
p_weekday <- ggplot(subset(activity_pattern_2, 
                           daytype == "Weekday", 
                           select= c(interval, average_steps)
                           ),
                    aes(x = interval,
                        y = average_steps
                        )
                    ) +
        geom_line(color = "blue",
                  size = 1
                  ) + 
        labs(title = "Average Steps per Interval for Weekdays", 
             y = "Average Steps",
             x = "Interval"
             ) +
        guides(color=guide_legend(title="Day Type")) +
        theme(legend.position="bottom")

p_weekend <- ggplot(subset(activity_pattern_2, 
                           daytype == "Weekend", 
                           select= c(interval, average_steps)
                           ),
                    aes(x = interval,
                        y = average_steps
                        )
                    ) +
        geom_line(color = "blue",
                  size = 1
                  ) + 
        labs(title = "Average Steps per Interval for Weekends", 
             y = "Average Steps",
             x = "Interval"
             ) +
        guides(color=guide_legend(title="Day Type")) +
        theme(legend.position="bottom")

grid.arrange(p_weekday, p_weekend, nrow=2)

```











